---
title: "Alpha div Captive vs Wild DMR"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---



```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(breakaway))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(grateful))
suppressPackageStartupMessages(library(wesanderson))
suppressPackageStartupMessages(library(DT))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(ggforce))
```



```{r constants}
```

```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccid')

metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = "")
)) %>% 
  left_join( # get the nb reads 
    asvs %>% select(sample, count) %>% 
    group_by(sample) %>% summarise(NbReads = sum(count), .groups = 'drop'))

taxonomy <- read_tsv("../data/taxonomy.tsv", col_types =  cols(
  asv = col_character(),
  kingdom = col_character(),
  phylum = col_character(),
  class = col_character(),
  order = col_character(),
  family = col_character(),
  genus = col_character(),
  species = col_character()
)) %>%
  naniar::replace_with_na_all(condition = ~.x %in%
                                c("uncultured bacterium", "Ambiguous_taxa","uncultured")) # replace all that is "uncultured bacterium", "uncultured"  & Ambiguous_taxa"with NA

# get palettes
pal <- wes_palette("Darjeeling1")
pal <- pal[c(2,3)]
pal <- rev(pal) # change order
pal2 <- pal
pal2[2:3] <- c('grey', pal2[c(2,3)])

#create a pseq object
asvsp <- asvs %>% select(asv, count, sample) %>% 
  left_join(metadata %>% select(sample, Asample)) %>% select(-sample) %>% # neeed another sample name, sometime problem with sample names starting with numbers only
  pivot_wider(names_from = 'Asample', values_from='count', values_fil= 0) %>%
  tibble::column_to_rownames('asv') %>% as.matrix()
taxonomyp <- taxonomy %>% tibble::column_to_rownames('asv') %>% as.matrix()
metadatap <- metadata %>% #mutate(Plate_No = as.factor(Plate_No)) 
            mutate_if(is.character,as.factor) %>% tibble::column_to_rownames('Asample')

pseq <- phyloseq(otu_table(asvsp, taxa_are_rows = TRUE), tax_table(taxonomyp), sample_data(metadatap))
rm(asvsp, taxonomyp, metadatap)
```


<p>&nbsp;</p>

# Observed Richness

```{r Observed richness, echo=FALSE, fig.cap='**Figure 1.** Observed alpha diversity (richness). Colour by Population.', message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
observed_c <- sample_richness(pseq)
observed_c <- summary(observed_c) %>% rename(Asample = sample_names) %>% left_join(metadata %>% select(Asample, Population, Plate_No))
observed_c %>% 
  ggplot(aes(x = Population, y = estimate, col = Population)) +
  geom_violin() +
  geom_sina() + 
  geom_violin() +
  ggforce::geom_sina() +
  ylab("Observed richness") +
  xlab("") +
  scale_colour_manual( values = pal) +
  scale_colour_manual( values = pal) +
      theme(axis.title.y = element_text(size =25, face = "bold"),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20),
        panel.background=element_rect(fill = 'white', colour='black'),
        strip.background=element_rect(fill = 'white', colour='white'),
        strip.text = element_text(face = "bold", size=15),
        panel.grid = element_blank(),
        legend.position = "none"
        )  +
  ylim(0, 700)
```


<p>&nbsp;</p>


# Breakaway Richness

```{r breakaway, echo=FALSE, fig.cap='**Figure 2.** Estimated alpha diversity (richness) with confidence interval with breakaway, sorted by estimate. Colour by Population.', message=FALSE, warning=FALSE, fig.height=8, fig.width=8,}
# Calculate estimated species richness breakaway
adiv_break <- pseq %>% breakaway()
adiv_break <- summary(adiv_break) %>% rename(Asample = sample_names) %>% left_join(metadata %>% select(Asample, Population, Plate_No))

adiv_break %>%
  mutate(Asample = fct_reorder(Asample, estimate)) %>%
  ggplot(aes(x = Asample, y = estimate,col = Population)) +
  geom_point() +
  scale_colour_manual( values = pal) +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  coord_flip() +
  xlab('Sample') + 
  ylab('Estimated richness with confidence interval') +
  ggtitle("Breakaway richness")
```
From `breakaway-vignette`: Don't freak out! Those are wide error bars, but nothing went wrong -- it's just really hard to estimate the true number of unknown species. `breakaway` was developed to deal with this, and to make sure that we account for that uncertainty when we do inference.


<p>&nbsp;</p>


```{r breakawayviolin, echo=FALSE, fig.cap='**Figure 3.** Estimated alpha diversity (richness) violins. Colour by Population.', message=FALSE, warning=FALSE, fig.height=5, fig.width=4}
p1 <- adiv_break %>%
  ggplot(aes(x = Population, y = estimate, col = Population)) +
  geom_violin() +
  geom_sina() + 
  geom_violin() +
  ggforce::geom_sina() +
  ylab("ASV richness")+
  xlab("") +
  scale_colour_manual( values = pal) +
      theme(axis.title.y = element_text(size =25, face = "bold"),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust= 1),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill = 'white', colour='black'),
        strip.background=element_rect(fill = 'white', colour='white'),
        strip.text = element_text(face = "bold", size=15),
        panel.grid = element_blank(),
        legend.position = "none"
        )  +
  ylim(0, 700)

p1
```

Ok that is very similar to Observed richness!

<p>&nbsp;</p>


Test the null-hypothesis that samples from the two populations did not differ in richness: fit a model with Population as fixed effect using the betta() function
```{r echo=FALSE}
bt_fixed <- betta(formula = estimate ~ Population, 
                      ses = error, data = adiv_break)
bt_fixed$table
```

Estimated Richness for Wild samples is 131 ASVS less than for Captive, with standard error of 10 of estimate. Suggest there is a large difference between Captive and Wild in Richness.

<p>&nbsp;</p>

Add plate_no as random factor using betta_random()
```{r echo=FALSE}
bt_fixed_random <-
  betta_random(formula = estimate ~ Population | Plate_No, 
               ses = error,  data = adiv_break)
bt_fixed_random$table
# get exact p-value pop
2 * (1 - pnorm(abs(bt_fixed_random$table["PopulationWild","Estimates"]/
                     bt_fixed_random$table["PopulationWild","Standard Errors"]))) # still 0
```


Similar result as not using random factor, slightly smaller error and estimate.

<p>&nbsp;</p>


`r figr('mean-rich', T, type = 'Table')`. Table of mean and sd richness wild and captive samples.
```{r mean-rich, echo=FALSE, warning=FALSE, cache=FALSE}
adiv_break %>% group_by(Population) %>%
  summarise(meanRich = mean(estimate),
            sdRich = sd(estimate)) %>%
  mutate_if(is.numeric, round, digits = 3) %>%
  kable()
```


<p>&nbsp;</p>

Now we would like to know in what taxa where Number of ASVs differ between wild and captive: 

```{r ASVrichness, fig.cap = '**Figure 4.** Number of ASVs per phylum. Samples are either "Captive" (orange) or "Wild" (green) by color and with boxplot show spread of raw data.', fig.width=12, fig.height = 6, message=FALSE, warning=FALSE, echo = FALSE}
# caculate top8 phyla with highest prevalence
top7phyla <- asvs %>% left_join(taxonomy) %>% select(sample, phylum) %>% unique() %>% 
  group_by(phylum) %>% summarise(Nsamples = n()) %>% arrange(desc(Nsamples)) %>% ungroup() %>%  
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    inner_join(taxonomy, by = 'asv') %>%
  group_by(phylum) %>% 
  summarise(reads = sum(count)) )%>% 
  top_n(7, Nsamples) %>% 
  arrange(desc(Nsamples), desc(reads)) %>% mutate(Order  = 1:nrow(.))

# plot the two most common
p2 <- asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(top7phyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  replace_na(list('topph' = 'Other phyla')) %>%
  group_by(sample, topph) %>%
  summarise(Nasvs = n()) %>%
  rename(phylum = topph) %>%
  ungroup() %>% 
  complete(sample, nesting(phylum), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  left_join(top7phyla) %>%
  replace_na(list('Nasvs' = 0)) %>%
  filter(Order < 3 ) %>%
  mutate(phylum =  fct_reorder(phylum, Order)) %>%
  ggplot(aes(x = phylum, y= Nasvs)) +
  labs(y = "N ASVs", x ="") +
  geom_boxplot(aes(x = phylum, y= Nasvs, fill=Population),
               position = position_dodge(width = .7)#, outlier.shape = NA
               )+
  scale_fill_manual(values = pal) +
  theme(legend.position = "none",
        axis.title.y = element_text(size =25, face = "bold"),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  guides(alpha =FALSE) 

#plot the less common
p3 <- asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(top7phyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  replace_na(list('topph' = 'Other phyla')) %>%
  group_by(sample, topph) %>%
   summarise(Nasvs = n()) %>%
  rename(phylum = topph) %>%
  ungroup() %>% 
  complete(sample, nesting(phylum), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  left_join(top7phyla) %>%
  replace_na(list('Order' = 11)) %>%
  filter(Order > 2 ) %>%
  mutate(phylum =  fct_reorder(phylum, Order)) %>%
  ggplot(aes(x = phylum, y= Nasvs)) +
  labs(y = "", x ="") +
  geom_boxplot(aes(x = phylum, y= Nasvs, fill=Population),
               position = position_dodge(width = .7) #, 
               #outlier.shape = NA
               )+
  scale_fill_manual(values = pal) +
  theme(legend.title = element_blank(),
        legend.position = c(0.8, 0.90),
        legend.direction = "horizontal",
        legend.text = element_text(size =20),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  guides(alpha =FALSE)

#comine to one plot
p4 <- ggarrange(p2+ ggtitle("A") , p3 +ggtitle("B") , nrow=1, ncol=2, align = "h", widths = c(3,8))
p4
```


So it looks like *Ruminococcaceae* is the reason for higher diversity within captive samples.

<p>&nbsp;</p>


```{r ASVrichnessfamFirmicutes, fig.cap = '**Figure 5.** Number of ASVs within families of Frimicutes. Samples are either "Captive" (orange) or "Wild" (green) by color and with boxplot show spread of raw data. Y-axis on log-10 scale.', fig.width=12, fig.height = 6, message=FALSE, warning=FALSE, echo = FALSE}

# pick out the more diverse families
firmfam<- asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  filter(phylum == "Firmicutes") %>%
  group_by(sample, family) %>%
  summarise(Nasvs = n()) %>%
  ungroup() %>% 
  complete(sample, nesting(family), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  replace_na(list('Nasvs' = 0)) %>%
  group_by(Population, family) %>% 
  summarise(meanNasvs = mean(Nasvs)) %>% ungroup() %>%
  filter(meanNasvs > 10) %>% select(family) %>% unique()

asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  filter(phylum == "Firmicutes") %>%
  group_by(sample, family) %>%
  summarise(Nasvs = n()) %>%
  ungroup() %>% 
  complete(sample, nesting(family), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  filter(family %in% firmfam$family) %>%
  ggplot(aes(x = family, y= Nasvs)) +
  labs(y = "N ASVs", x ="") +
  geom_boxplot(aes(x = family, y= Nasvs, fill=Population),
               position = position_dodge(width = .7)#, outlier.shape = NA
               )+
  scale_fill_manual(values = pal) +
  theme(legend.position = "none",
        axis.title.y = element_text(size =25, face = "bold"),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  guides(alpha =FALSE)
```


Not a very pretty plot, but we see that richness within *Ruminococcaceae* is much higher for captive than wild.

<p>&nbsp;</p>

```{r ASVrichnesscombined, fig.cap = '**Figure 6.** combined bplot for manuscript. A) ASV rrichness and B) Number of ASVs per phylum. Samples are either "Captive" (orange) or "Wild" (green) by color and with boxplot show spread of raw data.', fig.width=12, fig.height = 6, message=FALSE, warning=FALSE, echo = FALSE}
p5 <- ggarrange(p1 +ggtitle("A"), p2+ ggtitle("B"), p3 +ggtitle("C") , nrow=1, ncol=3, align = "h", widths = c(1,1,3))
p5
```


<p>&nbsp;</p>

`r figr('wilcoxNasvsP', T, type = 'Table')`. Table results wilcox test ASV richness of phyla.
```{r wilcoxNasvsP, echo=FALSE, message=FALSE, warning=FALSE}
# test N ASVS 
Pdata <- asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(top7phyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  replace_na(list('topph' = 'Other_phyla')) %>%
  group_by(sample, topph) %>%
  summarise(Nasvs = n()) %>%
  rename(phylum = topph) %>%
  ungroup() %>% 
  complete(sample, nesting(phylum), fill = list(Nasvs = 0)) %>% 
  #pivot_wider(names_from = 'phylum', values_from='Nasvs', values_fill= 0) %>%
  left_join(metadata %>% select(sample, Population))

vars = names(Pdata)[c(2:8)]


models <- Pdata  %>%
  split(.$phylum) %>% 
  map( ~ wilcox.test(Nasvs ~ Population, data = ., ) %>%  broom::tidy()) %>%
  bind_rows(.id = "phylum") %>%
  arrange(p.value) %>% 
  mutate(across(is.numeric, ~ round(., 6)),
         p.adjust = p.adjust(p.value, method="bonferroni")) # Bonferroni correction for that we did tests on 17 family groups 

tableNasvsP <- Pdata %>%
  group_by(phylum, Population) %>%
  summarise(meanNasvs = mean(Nasvs),
            sdNAsvs = sd(Nasvs)) %>% ungroup() %>%
  pivot_wider(names_from = "Population", values_from = c("meanNasvs", "sdNAsvs")) %>%
  mutate_if(is.numeric,
                     round,
                     digits = 1) %>%
  left_join(
models  %>%
  select(phylum, p.value, p.adjust) %>%
  mutate_if(is.numeric,
                     round,
                     digits = 3) %>%
  mutate(p.value = as.character(p.value),
  p.value = case_when(p.value == 0 ~ "<0.001",
                      is.na(p.value) ~ "", 
                      TRUE ~ p.value),
  p.adjust = as.character(p.adjust),
  p.adjust = case_when(p.adjust == "0" ~ "<0.001",
                       TRUE ~ p.adjust)) 
) %>%
  left_join(top7phyla %>% select(phylum, Order)) %>% 
  arrange(Order) %>% 
  select(-Order)

# write
#tableNasvsP %>%
#  write.table("../tables/NasvsPhylymWilcox.txt")


tableNasvsP %>% kable()

```


<p>&nbsp;</p>
 
# R and packages


This analysis was run with `r R.version$version.string`. Versions of packages used can be found in `r figr('cite-packages', T, type = 'Table')`.

```{r cite-packages, echo=FALSE, warning=FALSE, cache=FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```


<p>&nbsp;</p>

# References
