---
title: "ANCOMBC Wild vs Captive "
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---


```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(phyloseq))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(microbiome))
suppressPackageStartupMessages(library(ANCOMBC))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(grateful))
```

```{r constants}
```

```{r read-data, message=FALSE, include=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccid')
metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  massindex = col_integer(),
  SampleDate = col_date(format = "")
)) 
taxonomy <- read_tsv("../data/taxonomy.tsv", col_types =  cols(
  asv = col_character(),
  kingdom = col_character(),
  phylum = col_character(),
  class = col_character(),
  order = col_character(),
  family = col_character(),
  genus = col_character(),
  species = col_character()
))

# rename unclassified
taxonomy  <- taxonomy %>%
  naniar::replace_with_na_all(condition = ~.x %in%
                                c("uncultured bacterium", "Ambiguous_taxa", "uncultured")) %>%
  mutate(
    phylum = ifelse(is.na(phylum), sprintf("%s unclassified", kingdom), phylum),
    class = ifelse(is.na(class), sprintf("%s unclassified", str_remove(phylum, "unclassified")), class),
    order = ifelse(is.na(order), sprintf("%s unclassified", str_remove(class, "unclassified")), order),
    family = ifelse(is.na(family), sprintf("%s unclassified", str_remove(order, "unclassified")), family),
    genus = ifelse(is.na(genus), sprintf("%s unclassified", str_remove(family, "unclassified")), genus),
    species = ifelse(is.na(species), sprintf("%s unclassified", str_remove(genus, "unclassified")), species))
```

# Introduction

testing for differential abundance testing between wild and captive samples with ANCOM (https://www.nature.com/articles/s41467-020-17041-7)


# Results


## ASVs

```{r make-phyloseq-object, include = FALSE}
str(asvs)

asvsp <- asvs %>% left_join(metadata %>% select(sample, Asample)) %>% # join in as sometimes samplename gets strange if only numbers
  select(asv, count, Asample) %>%
  pivot_wider(names_from = 'Asample', values_from='count', values_fil= 0) %>%
  tibble::column_to_rownames('asv') %>% as.matrix()

taxonomyp <- taxonomy %>% tibble::column_to_rownames('asv') %>% as.matrix()
metadatap <- metadata %>% tibble::column_to_rownames('Asample')

OTU = otu_table(asvsp, taxa_are_rows = TRUE)
TAX = tax_table(taxonomyp)
samples = sample_data(metadatap)
  
pseq <- phyloseq(OTU, TAX, samples) # make pseq object
print(pseq)

rm(OTU, TAX, samples, taxonomyp, metadatap)
```



```{r run-ancombc, include = FALSE}
# run ancombc ASV - same result if singletons were filtered out before running
out <- ancombc(phyloseq = pseq, formula = "Population", 
              p_adj_method = "BH", zero_cut = 0.90, lib_cut = 1000, 
              neg_lb = TRUE, tol = 1e-5, 
              max_iter = 100, conserve = TRUE, alpha = 0.05, global = FALSE)

res <- out$res

df <- res$beta %>% rownames_to_column("asv") %>% rename(Coef = PopulationWild) %>%
  left_join(
res$se  %>% rownames_to_column("asv") %>% rename(se = PopulationWild)) %>%
  left_join(
res$W %>% rownames_to_column("asv") %>% rename(Testst = PopulationWild)) %>%
  left_join(
res$p_val %>% rownames_to_column("asv") %>% rename(Pval = PopulationWild)) %>%
  left_join(
res$q %>% rownames_to_column("asv") %>% rename(AdjPval = PopulationWild)) %>%
  left_join(
res$diff_abn %>% rownames_to_column("asv") %>% rename(DiffAb = PopulationWild))

# write table to plot with loading scores in betadiv script that is later used in "Betadiv.Rmd"
#df %>% 
#   mutate(overrep_inA= case_when(DiffAb ==TRUE & AdjPval < 0.05  & Coef > 0 ~ "Wild",
#                           DiffAb ==TRUE & AdjPval < 0.05 & Coef < 0 ~ "Captive",
#                           TRUE ~ "Notdiff"), 
#          .before =1) %>% filter(overrep_inA !="Notdiff") %>%
# write_tsv("../data/TableS2.tsv")
```



<p>&nbsp;</p>


```{r plot-ancombc-asv,  fig.cap='**Figure 1.** Plot of ASVS identidied as differentiallt abundant wild and captive samples.',fig.height=15, fig.width=12, message=FALSE}
df %>% 
  left_join(taxonomy) %>%
  left_join(df  %>% left_join(taxonomy) %>% select(phylum) %>% unique() %>% arrange(phylum) %>% mutate(Order  = 1:nrow(.))) %>%
  mutate(family = fct_reorder(family, Order)) %>% 
  filter(DiffAb ==TRUE & AdjPval < 0.05) %>%
  ggplot(aes(x = Coef, y = family, colour = phylum)) + 
      geom_point(alpha = 0.5 , size = 2) +
      geom_errorbar(aes(xmin = Coef -se, xmax = Coef + se),
                    width = 0, lwd = 1, alpha = 0.5) +
      scale_x_continuous(name="\nLog fold change") + 
      ylab("Family \n")  +
  geom_vline(xintercept = 0, linetype="dotted", size=1.5) +
      theme(axis.text = element_text(size=20),
            axis.title = element_text(size=22), 
            legend.text = element_text(size=20),
            legend.title = element_text(size=22),
            panel.grid.major = element_blank(),
            panel.grid.minor = element_blank()) 
```


<p>&nbsp;</p>


`r figr('sum-asvs', T, type = 'Table')` N ASVs over represented in wild (Coef > 0) or captive ( Coef < 0) samples with padj < 0.05 by ancombc.
```{r sum-asvs, echo=FALSE}
df %>%
  mutate(overrep_inA= case_when(DiffAb ==TRUE & AdjPval < 0.05  & Coef > 0 ~ "Wild",
                           DiffAb ==TRUE & AdjPval < 0.05 & Coef < 0 ~ "Captive",
                           TRUE ~ "Notdiff"),
         .before  =1) %>% 
  rename('over represented in' = overrep_inA) %>% 
  group_by(`over represented in`) %>% summarise(n =n()) %>% kable()
```


<p>&nbsp;</p>


# R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in
`r figr('cite-packages', T, type = 'Table')`.


```{r cite-packages, cache = FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```



<p>&nbsp;</p>

# References