---
title: "Captive VS Wild DMR beta diversity"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---


```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(wesanderson))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(grateful))
suppressPackageStartupMessages(library(compositions))
```

```{r constants}
```

```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccid')

metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = ""),
  massindex = col_double()
)) %>% 
  left_join( # get the nb reads 
    asvs %>% select(sample, count) %>% 
    group_by(sample) %>% summarise(NbReads = sum(count), .groups = 'drop'))

taxonomy <- read_tsv("../data/taxonomy.tsv", col_types =  cols(
  asv = col_character(),
  kingdom = col_character(),
  phylum = col_character(),
  class = col_character(),
  order = col_character(),
  family = col_character(),
  genus = col_character(),
  species = col_character()
)) %>%
  naniar::replace_with_na_all(condition = ~.x %in%
                                c("uncultured bacterium", "Ambiguous_taxa","uncultured")) # replace all that is "uncultured bacterium", "uncultured"  & Ambiguous_taxa"with NA

# get ancomb asvs data
ancomb <- read_tsv("../data/TableS2.tsv") # created in ancom.Rmd script

# get palette
pal <- wes_palette("Darjeeling1")
pal <- pal[c(2,3)]
pal <- rev(pal) # change order
pal2 <- pal
pal2[2:3] <- c('grey', pal2[c(2,3)])
```


# PCA CLR-transformed counts

```{r clr1, include=FALSE}
# make a matrix with row sample and column ASV with counts
m1 <- asvs  %>% 
  select(asv, sample) %>%
  complete(asv, sample) %>% 
  left_join(asvs %>% select(asv, count, sample)) %>%
  mutate( count = replace_na(count, 0),
          count = count + 0.65) %>% # add psuedo count of 0.65
  pivot_wider(names_from = 'asv', values_from='count') %>% 
  tibble::column_to_rownames('sample') %>% as.matrix()

### clr-transformation of the counts ###
clr <- t(apply(t(m1), 2, compositions::clr))

rm(m1)
```



```{r PCAclr1, message=FALSE, warning=FALSE, include=FALSE}
### Perform the PCA ###
pca <- vegan::rda(clr)

pca.samples <- pca$CA$u %>% data.frame() %>% tibble::rownames_to_column('sample')
pca.asv    <- pca$CA$v %>% data.frame() %>% tibble::rownames_to_column('asv')
pca.eigs    <- pca$CA$eig %>% data.frame() %>% tibble::rownames_to_column('pc') %>%
    rename(eigval = 2) %>%
  mutate(propexpl = eigval/sum(eigval))

# Calculate the variance explained by PC1, PC2 & PC3
d.mvar <- sum(pca$sdev^2) # total variance
PC1 <- paste("PC1 ","(",round(pca.eigs[1,3]*100,1),"%",")", sep="" )
PC2 <- paste("PC2 ","(",round(pca.eigs[2,3]*100,1),"%",")",sep="")
PC3 <- paste("PC3 ","(",round(pca.eigs[3,3]*100,1),"%",")",sep="")

```


<p>&nbsp;</p>


```{r clrplot1, echo=FALSE, message=FALSE,  warning=FALSE, echo=FALSE,  fig.height=6, fig.width=12,fig.cap='**Fig 1.** PCA on clr transformed counts with pseudocount 0.65. Colour by Population.'}
#Add metadata to pca df
df1 <- pca.samples  %>% select(sample, PC1, PC2, PC3) %>% 
  left_join(metadata)

PCACLR1 <- df1 %>% 
  ggplot( aes(x = PC1, y = PC2)) + 
  geom_point(aes(col = Population), size = 2) +  
  scale_colour_manual( values = pal) +
  ylab(PC2) +
  xlab(PC1) +
 theme( axis.title = element_text(size =25, face = "bold"),
        axis.text = element_text(size =20),
        panel.background = element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        legend.position = "top",
        legend.key=element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size =20),
        panel.border = element_rect(colour = "black", fill=NA))+
  guides(alpha =FALSE, fill =FALSE)
PCACLR1    
```


*Note*: also tested pseudo count 1/Library size, got similar results with minor differences in % explained PC1 and PC2 - 0.65 had slightly better explanation PC1.


<p>&nbsp;</p>


```{r clrplot2, echo=FALSE, message=FALSE, warning=FALSE, echo=FALSE, fig.height=6, fig.width=12,fig.cap='**Fig 2.** PCA on clr transformed counts with pseudo count 0.65, PC1 and PC3. Colour by Population.' }

# plot by population
PCACLR1b <- df1 %>% 
  ggplot( aes(x = PC1,y = PC3)) + 
  geom_point(aes(col=Population), size = 2) +  
  scale_colour_manual( values = pal) +
  ylab(PC3) +
  xlab(PC1) +
 theme( axis.title = element_text(size =25, face = "bold"),
        axis.text = element_text(size =20),
        panel.background = element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        legend.position = "top",
        legend.key=element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size =20),
        panel.border = element_rect(colour = "black", fill=NA))+
  guides(alpha =FALSE, fill =FALSE)

PCACLR1b
```


<p>&nbsp;</p>


```{r clrplotPC1P2PC3, echo=FALSE, fig.cap = '**Fig 3.** PCA on clr transformed counts with pseudocount 0.65, A) PC1 and PC2, B) PC1 and PC3 Colour by Population.', message=FALSE, warning=FALSE, fig.width=12, fig.height=6}

ggarrange(PCACLR1 + ggtitle("A"),
          PCACLR1b + ggtitle("B"),
          common.legend = TRUE, nrow= 1)

```


<p>&nbsp;</p>


```{r PC1density, echo=FALSE, fig.cap = '**Fig 3b.** Density of PC1 values wild and captive samples. Another way to plot separation.', message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
df1 %>% ggplot( aes(x=PC1, fill=Population)) +
  geom_density() +
  scale_fill_manual( values = pal) +
  theme(axis.title=element_text(size=18,face="bold"),
        axis.text = element_text(size=14),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        legend.position = "top",
        legend.title = element_blank(),
        legend.text = element_text( size=14, face="bold")) +
  xlim(-0.17, 0.16)
```

<p>&nbsp;</p>


**Permanova Wild vs Captive on clrcounts, with Plate number nested strata argument:**
```{r permanova-adonis-clr, echo=FALSE, message=FALSE, warning=FALSE}
metadf <- metadata %>% arrange(sample) %>% tibble::column_to_rownames('sample') %>% data.frame()
clrdist <- clr %>% vegdist(method = "euclidean") 

#adonis2
perm <- how(nperm = 9999)
setBlocks(perm) <- with(metadf, Plate_No)
set.seed(2000)
adonis2(clrdist ~ NbReads+ Population, data = metadf, permutations = perm, by = "margin")

# betadisp
set.seed(2000)
betadisp <- betadisper(clrdist, metadf$Population)
anova(betadisp) 
#plot(betadisp)
```


<p>&nbsp;</p>

## Loading scores PC1 CLR
```{r loadingCLR }
#| echo = FALSE, 
#| fig.width = 12, fig.height = 8, 
#| fig.cap = '**Fig 4.** Identifying taxa more common in wild and captive animals from loading scores of ASVS driving dissimilarity on PC1 on clr transformed counts, A) Loading scores PC1 sorted for all ASV in dataset, the top  50 (1%) with the largest PC1 loadings are filled in green blue and correspond to ASVS characterizing wild gut microbiota while the bottom  50 (1%) with the lowest PC1 are filled in orange and correspond to ASVS characterizing captive gut microbiota of Damaraland mole-rats. B) N of ASVS belonging to phylum as either associated with "captive" or "wild".',
#|  message = FALSE, 
#|  warning = FALSE

pca.asv <- pca.asv %>% arrange(PC1)  %>%  mutate(ASV = 1:nrow(.))
maxasv <- max(pca.asv$ASV)

loadings <- 
  pca.asv %>% 
  select(asv, PC1) %>% 
  left_join(taxonomy) %>% # join in taxa
  mutate(
    phylum = ifelse(is.na(phylum), sprintf("%s unclassified", kingdom), phylum),
    class = ifelse(is.na(class), sprintf("%s unclassified", str_remove(phylum, "unclassified")), class),
    order = ifelse(is.na(order), sprintf("%s unclassified", str_remove(class, "unclassified")), order),
    family = ifelse(is.na(family), sprintf("%s unclassified", str_remove(order, "unclassified")), family),
    genus = ifelse(is.na(genus), sprintf("%s unclassified", str_remove(family, "unclassified")), genus),
    species = ifelse(is.na(species), sprintf("%s unclassified", str_remove(genus, "unclassified")), species)) %>%
  arrange(PC1) %>% 
  mutate(ASV = 1:nrow(.),
         col = case_when( ASV <= round(maxasv * 0.02)  ~ "Captive", # Mark bar to highlight in loading plot bottom and top ~2% = 101 ASVs each direction 
                          ASV > (round(maxasv * 0.98)) ~ "Wild", 
                          TRUE ~ "NA"),
         .before = 1) 

load <- loadings %>%
  ggplot(aes(x = ASV, y = PC1, fill = col)) +
  geom_col(width = 1) + 
  scale_fill_manual(breaks = c("Captive", "", "Wild" ), values = pal2) +
  ylab("Loading score PC1") +
  guides(fill = guide_legend(title = "Overrepresented in", title.position = "left")) +
  scale_x_continuous(limits = c(-10 ,maxasv + 10), expand = c(0, 0)) +
             scale_y_continuous("Loading score PC1", 
                                sec.axis = sec_axis(~ ., name = "Loading score PC1")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        legend.position = "top",
        legend.text = element_text(size=15),
        legend.title = element_text(size=15),
        axis.title = element_text(size=12),
        axis.text.y = element_text(size=12)) 

barsload <- loadings  %>%
  filter(col != "NA") %>% # pick out the one highlighted in plot A
  ggplot(aes(x = phylum)) +
  geom_bar() +
  facet_wrap(~col) +
  coord_flip() +
  theme(legend.position = "top",
        axis.title.x = element_text(size=15),
        axis.text.y = element_text(size=15),
        axis.text.x = element_text(size=15),
        strip.text = element_text(size=20),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  xlab("") +
  ylab("N ASVs")

clrload <- ggarrange(load, barsload, ncol=1, nrow =2,
          labels = c("A", "B"))
clrload
```


<p>&nbsp;</p>



```{r loadingCLRfam, echo=FALSE, fig.cap='**Fig 5.** A) Loading scores PC1 sorted for all ASV in dataset, the top  2 % with the largest PC1 loadings are filled in green blue and correspond to ASVS characterizing wild gut microbiota while the bottom  2% with the lowest PC1 are filled in orange and correspond to ASVS characterizing captive gut microbiota of Damaraland mole-rats. B) Loading scores of ASVs of the families with the ASVs in the top (1%) largest and the bottom (1%) PC1 loadings  combined as families.', fig.height=11, fig.width=10, message=FALSE, warnings=FALSE }
loadings2 <- loadings  %>%
  rename(overrep_inLoading = col) %>%
  left_join(ancomb %>% select(asv, DiffAb, Coef, AdjPval, overrep_inA)) 

signasvs  <- loadings2 %>%
  filter(!is.na(overrep_inA)) %>% # to get N asvs identified as sign abundant
  mutate(Pfam = paste(phylum, family, sep= " "), .before =1) %>%
  arrange(Pfam) %>% 
  group_by(Pfam, overrep_inA) %>%
  summarise( n = n())  %>%
  ungroup() %>%
  complete(Pfam, overrep_inA) %>% 
  mutate(n = replace_na(n, 0),
         ypos = case_when(overrep_inA == "Captive" ~ -1.4,
                          overrep_inA == "Wild" ~ 2.1,
                          ))

#signasvs %>% summarise(n = sum(n)) # double check ads up  to 690

pfamloadtop <- loadings2  %>%
  filter(overrep_inLoading != "NA") %>%
  mutate(Pfam = paste(phylum, family, sep= " "), .before =1)  %>% select(Pfam) %>%
  unique() %>% pull() # 30

labels <- signasvs  %>% filter(Pfam %in% pfamloadtop )

#labels %>% filter(Pfam %in% pfamloadtop ) %>% summarise( N = sum(n)) # 620 ASVs within the families of the top and tail 2% loading scores


pfam <- loadings2  %>%
  mutate(Pfam = paste(phylum, family, sep= " "), .before =1) %>%
  filter(overrep_inLoading != "NA" ) %>% # only plot the loading scores within top or tail 2 % - too much niose otherwise
  group_by(Pfam, overrep_inLoading) %>%
  summarise(PC1 = sum(PC1)) %>%
  #filter(overrep_inA == "Captive" & PC1 > 0) # check both go same dir
  #filter(overrep_inA == "Wild" & PC1 < 0) # check both go same dir
  ggplot(aes(x = Pfam, y= PC1)) +
  geom_col(aes(fill = overrep_inLoading), position = "stack") +
  geom_label(data = labels, aes(label = n, y = ypos)) +
  scale_fill_manual(values = pal) +
  coord_flip() +
  theme(legend.position = "top",
        axis.title.x = element_text(size=15),
        axis.text.y = element_text(size=15),
        axis.text.x = element_text(size=15),
        strip.text = element_text(size=20),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  guides(fill = guide_legend(title = "Overrepresented in", title.position = "left")) +
  xlab("") +
  ylab("Loading score")

ggarrange( load + xlab(""),
           pfam + theme(legend.position = "none"),
           ncol=1, nrow =2, heights = c( 0.7, 2),
          labels = c("A", "B"))

```


  
<p>&nbsp;</p>


# R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in
`r figr('cite-packages', T, type = 'Table')`.


```{r cite-packages, cache = FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```



<p>&nbsp;</p>

# References