---
title: "Relative abundances DMR Wild vs Captive"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---


```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(wesanderson))
suppressPackageStartupMessages(library(grateful))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(sjPlot))
suppressPackageStartupMessages(library(VennDiagram))
```

```{r constants}
```


```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccid')

metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = ""),
  massindex = col_double()
)) %>% 
  left_join( # get the nb reads 
    asvs %>% select(sample, count) %>% 
    group_by(sample) %>% summarise(NbReads = sum(count), .groups = 'drop'))

taxonomy <- read_tsv("../data/taxonomy.tsv", col_types =  cols(
  asv = col_character(),
  kingdom = col_character(),
  phylum = col_character(),
  class = col_character(),
  order = col_character(),
  family = col_character(),
  genus = col_character(),
  species = col_character()
))  %>%
  naniar::replace_with_na_all(condition = ~.x %in%
                                c("uncultured bacterium", "Ambiguous_taxa", "uncultured")) # replace with NA as unassigned

# get palette
pal <- wes_palette("Darjeeling1")
pal <- pal[c(2,3)]
pal <- rev(pal) # change order

```

# Introduction

In this study, we are comparing fecal samples from captive individuals with wild caught animals. 

<p>&nbsp;</p>


# Results


`r figr('unassigned_taxa', T, type = 'Table')`. Table of assigned taxa of ASVs in wild and captive samples.

```{r assigned_taxa, echo=FALSE, message=FALSE}
assign_taxa <- data.frame(Population=character(),kingdom = double(),
                   phylum = double(), class = double(), 
                   order = double(), family = double(), 
                   genus = double(), species = double())
#asvs %>% left_join(taxonomy) %>% 
#  select(asv, kingdom) %>% unique() %>% filter(!is.na(kingdom)) %>%  summarise(n = n()) %>% pull() # all assigned to kingdom

# phylum
c_phyl <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, phylum) %>% unique() %>% filter(!is.na(phylum)) %>%  summarise(n = n()) %>% pull() /
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, phylum) %>% unique()  %>%  summarise(n = n()) %>% pull() 
w_phyl <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, phylum) %>% unique() %>% filter(!is.na(phylum)) %>%  summarise(n = n()) %>% pull()/
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, phylum) %>% unique()  %>%  summarise(n = n()) %>% pull()

# class
c_class <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, class) %>% unique() %>% filter(!is.na(class)) %>%  summarise(n = n()) %>% pull() /
   asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, class) %>% unique()  %>%  summarise(n = n()) %>% pull() 
w_class <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, class) %>% unique() %>% filter(!is.na(class)) %>%  summarise(n = n()) %>% pull() /
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, class) %>% unique() %>%  summarise(n = n()) %>% pull()

# order
c_ord <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, order) %>% unique() %>% filter(!is.na(order)) %>%  summarise(n = n()) %>% pull() / 
   asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, order) %>% unique() %>%  summarise(n = n()) %>% pull()
w_ord <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, order) %>% unique() %>% filter(!is.na(order)) %>%  summarise(n = n()) %>% pull()/
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, order) %>% unique() %>%  summarise(n = n()) %>% pull()

# family
c_fam <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, family) %>% unique() %>% filter(!is.na(family)) %>%  summarise(n = n()) %>% pull() / 
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, family) %>% unique()  %>%  summarise(n = n()) %>% pull() 
w_fam <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, family) %>% unique() %>% filter(!is.na(family)) %>%  summarise(n = n()) %>% pull() / 
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, family) %>% unique() %>% summarise(n = n()) %>% pull()

# genus
c_gen <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, genus) %>% unique() %>% filter(!is.na(genus)) %>%  summarise(n = n()) %>% pull() /
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, genus) %>% unique() %>%  summarise(n = n()) %>% pull()
w_gen <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, genus) %>% unique() %>% filter(!is.na(genus)) %>%  summarise(n = n()) %>% pull()  / 
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, genus) %>% unique() %>%  summarise(n = n()) %>% pull() 

# species
c_sp <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, species) %>% unique() %>% filter(!is.na(species)) %>%  summarise(n = n()) %>% pull() / 
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Captive") %>%
  select(asv, genus) %>% unique() %>%  summarise(n = n()) %>% pull() 
w_sp <- asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, species) %>% unique() %>% filter(!is.na(species)) %>%  summarise(n = n()) %>% pull()  /
  asvs %>% left_join(taxonomy) %>% left_join(metadata %>% select(sample, Population)) %>% filter( Population == "Wild") %>%
  select(asv, genus) %>% unique() %>%  summarise(n = n()) %>% pull()

dt <- assign_taxa %>% 
  #add_row(Population = "Full dataset", kingdom = 0,
  #        phylum = a_phyl, class = a_class, order = a_ord,
  #        family = a_fam, genus = a_gen, species = a_sp) %>%
  add_row(Population = "Captive", kingdom = 0,
          phylum = c_phyl, class = c_class, order = c_ord,
          family = c_fam, genus = c_gen, species = c_sp) %>%
  add_row(Population = "Wild", kingdom = 0,
          phylum = w_phyl, class = w_class, order = w_ord,
          family = w_fam, genus = w_gen, species = w_sp) %>% 
  mutate( kingdom = scales::percent(kingdom),
          phylum = scales::percent(phylum),
          class = scales::percent(class),
         order = scales::percent(order),
          family = scales::percent(family),
          genus = scales::percent(genus),
          species = scales::percent(species)) %>%
  rename(domain = kingdom)
dt %>% kable()

# write table
#dt %>% write.table(file ="../tables/unassignedtaxa.txt", row.names = F)

rm(c_class, c_fam, c_gen, c_ord, c_phyl, c_sp, w_class, w_fam, w_gen, w_ord, w_phyl, w_sp,assign_taxa, dt)
```


```{r rename_unassigned_taxa, echo=FALSE, message=FALSE}
# rename unclassified
taxonomy  <- taxonomy %>%
  mutate(
    phylum = ifelse(is.na(phylum), sprintf("%s unclassified", kingdom), phylum),
    class = ifelse(is.na(class), sprintf("%s unclassified", str_remove(phylum, "unclassified")), class),
    order = ifelse(is.na(order), sprintf("%s unclassified", str_remove(class, "unclassified")), order),
    family = ifelse(is.na(family), sprintf("%s unclassified", str_remove(order, "unclassified")), family),
    genus = ifelse(is.na(genus), sprintf("%s unclassified", str_remove(family, "unclassified")), genus),
    species = ifelse(is.na(species), sprintf("%s unclassified", str_remove(genus, "unclassified")), species))

```


<p>&nbsp;</p>



### Compare relative abundance phylum level


```{r TaxonomyplotWildvsCapt, echo=FALSE, fig.cap='**Figure 1a.** Taxonomyplot top 8 phyla faeces samples.Samples are either "Captive" (Mole-rat lab) or "Wild" (Kuruman River Reserve population)', fig.height=8, fig.width=12, message=FALSE, cache=FALSE}
# caculate top8 phyla with highest prevalence
topphyla <- asvs %>% left_join(taxonomy) %>% select(sample, phylum) %>% unique() %>% 
  group_by(phylum) %>% summarise(Nsamples = n()) %>% arrange(desc(Nsamples)) %>% ungroup() %>%  
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    inner_join(taxonomy, by = 'asv') %>%
  group_by(phylum) %>% 
  summarise(reads = sum(count)) )%>% 
  top_n(7, Nsamples) %>% 
  arrange(desc(Nsamples), desc(reads)) %>% mutate(Order  = 1:nrow(.))

# caculate top10 class
topclass <- asvs %>% 
  left_join(taxonomy) %>% select(sample, phylum, class) %>% unique() %>% 
  group_by(phylum, class) %>% summarise(Nsamples = n()) %>% arrange(desc(Nsamples)) %>% ungroup() %>%  
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    inner_join(taxonomy, by = 'asv') %>%
  group_by(class) %>% 
  summarise(reads = sum(count)) )%>% 
  top_n(10, Nsamples) %>% 
  arrange(desc(Nsamples), desc(reads)) %>% mutate(Order  = 1:nrow(.))

ggarrange(
  asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(topphyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  group_by(topph, sample) %>%
  replace_na(list('topph' = 'Other phyla')) %>%
  summarise(relab = sum(relab)) %>%
  left_join(metadata %>% select(sample, Population)) %>%
  ggplot(aes(x = sample, y = relab, fill = topph)) +
  labs(y = "Relative abundance") +
  geom_col(position = 'fill') +
  labs(fill = "Phylum") + 
  facet_grid(cols = vars(Population),  scales="free") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x =element_blank(),
        axis.line.x=element_blank()),
  asvs %>% 
  inner_join(taxonomy, by = 'asv') %>%
  left_join(topclass %>% transmute(class, topph = class), by = 'class') %>%
  group_by(topph, sample) %>%
  replace_na(list('topph' = 'Other Classes')) %>%
  summarise(relab = sum(relab)) %>%
  left_join(metadata %>% select(sample, Population)) %>%
  ggplot(aes(x = sample, y = relab, fill = topph)) +
  labs(y = "Relative abundance") +
  geom_col(position = 'fill') +
  labs(fill = "Class") + 
  facet_grid(cols = vars(Population),  scales="free") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x =element_blank(),
        axis.line.x=element_blank()),
  labels = c("A", "B"),
          ncol = 1, nrow = 2)

```


<p>&nbsp;</p>


`r figr('unassigned_taxa', T, type = 'Table')`. Number of reads  and sample of all phylum within dataset.
```{r table1phyl, echo=FALSE, message=FALSE, warnings=FALSE, echo = FALSE}
phylatable <- asvs %>% left_join(taxonomy) %>% select(sample, phylum) %>% unique() %>% 
  group_by(phylum) %>% summarise(Nsamples = n()) %>% arrange(desc(Nsamples)) %>% ungroup() %>%  
  left_join(
  asvs %>% left_join(metadata %>% select(sample, Population)) %>% left_join(taxonomy) %>% select(sample,Population, phylum) %>% unique() %>% 
  group_by(phylum, Population) %>% 
  summarise(Nsamples = n()) %>% pivot_wider(names_from = Population, values_from = Nsamples)) %>%
  rename(NsamplesWild = Wild,  NsamplesCaptive = Captive) %>%
  left_join(
  asvs %>% filter(!is.na(count)) %>% left_join(metadata %>% select(sample, Population)) %>%
    inner_join(taxonomy, by = 'asv') %>%
  group_by(phylum, Population) %>% 
  summarise(reads = sum(count)) %>% pivot_wider(names_from = Population, values_from = reads) %>%
  rename(ReadsWild = Wild, ReadsCaptive = Captive)) %>%
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    inner_join(taxonomy, by = 'asv') %>%
  group_by(sample, phylum) %>%
    summarise(count = sum(count),
              relab = sum(relab) )%>%
    group_by(phylum) %>%
  summarise(reads = sum(count),
            meanrelab = mean(relab),
            maxrelab = max(relab),
            minrelab = min(relab),
            sdrelab = sd(relab) ) 
  ) %>%  arrange(desc(Nsamples))

phylatable %>% kable(digits = 4)
```



<p>&nbsp;</p>



```{r relabsbox2, fig.cap = '**Figure 2.** Relative abundances of phylum with 100 % prevalence. Samples are either "Captive" (orange) or "Wild" (green) by color and with boxplot show spread of raw data.', fig.width=12, fig.height = 6, message=FALSE, warning=FALSE, echo = FALSE}
plot2 <-asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(topphyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  replace_na(list('topph' = 'Other phyla')) %>%
  group_by(sample, topph) %>%
  summarise(relab = sum(relab)) %>%
  rename(phylum = topph) %>%
  ungroup() %>% 
  complete(sample, nesting(phylum), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  left_join(topphyla) %>%
  replace_na(list('meanrelab' = 0)) %>%
  filter(Order < 3 ) %>%
  mutate(phylum =  fct_reorder(phylum, Order)) %>%
  ggplot(aes(x = phylum, y= relab)) +
  labs(y = "Relative Abundance", x ="") +
  geom_boxplot(aes(x = phylum, y= relab, fill=Population),
               position = position_dodge(width = .7)#, outlier.shape = NA
               )+
  scale_fill_manual(values = pal) +
  theme(legend.position = "none",
        axis.title.y = element_text(size =25, face = "bold"),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  ylim(0, 0.9) +
  guides(alpha =FALSE) + ggtitle("A")

plot1 <- asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(topphyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  replace_na(list('topph' = 'Other phyla')) %>%
  group_by(sample, topph) %>%
  summarise(relab = sum(relab)) %>%
  rename(phylum = topph) %>%
  ungroup() %>% 
  complete(sample, nesting(phylum), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  left_join(topphyla) %>%
  replace_na(list('Order' = 11)) %>%
  filter(Order > 2 ) %>%
  mutate(phylum =  fct_reorder(phylum, Order)) %>%
  ggplot(aes(x = phylum, y= relab)) +
  labs(y = "", x ="") +
  ggtitle("B")+
  geom_boxplot(aes(x = phylum, y= relab, fill=Population),
               position = position_dodge(width = .7) #, 
               #outlier.shape = NA
               )+
  scale_fill_manual(values = pal) +
  theme(legend.title = element_blank(),
        legend.position = c(0.8, 0.90),
        legend.direction = "horizontal",
        legend.text = element_text(size =20),
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  ylim(0, 0.15) +
  guides(alpha =FALSE)
ggarrange(plot2 , plot1, nrow=1, ncol=2, align = "h", widths = c(3,8))
```


<p>&nbsp;</p>



```{r relabsboxALL, echo=FALSE, fig.cap='**Figure 3.** Relative abundances of A) two most abundant phylum,  B) next 6 most common phyla (100 % prevalence) and C) Rest of all phyla ("Other" in fig B), of low relative abundance** Samples are either "Captive" (orange) or "Wild" (green) by colour and with boxplot show spread of raw data. Note the log scale relative abundance in plot C as these phyla in general had very low relative abundance.', fig.height=12, fig.width=12, warning=FALSE, message=FALSE}

plot3 <- asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  group_by(sample, phylum) %>%
  summarise(relab = sum(relab)) %>%
  ungroup() %>% 
  complete(sample, nesting(phylum), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  left_join(topphyla) %>%
  filter(is.na(Nsamples)) %>% # get phyla not in top 8
  ggplot(aes(x = phylum, y= relab)) +
  labs(y = "Log Relative Abundance", x ="") +
  ggtitle("C") +
  geom_boxplot(aes(x = phylum, y= log(relab), fill=Population),
               position = position_dodge(width = .7) )+
  scale_fill_manual(values = pal) +
  theme(axis.title.y = element_text(size =20, face = "bold"),
        legend.position = "none",
        axis.text.y = element_text(size =20),
        axis.text.x = element_text(size =20, angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key=element_blank(),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white')) +
  guides(alpha =FALSE)

ggarrange(ggarrange(plot2 , plot1,nrow=1, ncol=2, align = "h", widths = c(3,8)), plot3, nrow=2, ncol=1)
```


<p>&nbsp;</p>



### Compare relative abundance family level

```{r relabsboxplotfam, fig.width=10, fig.height = 8, fig.cap = '**Figure 4.** Relative abundances of the 12 families with highest prevalence (all samples).Samples are either "Captive" (orange) or "Wild" (green) by color and with boxplot show spread of raw data.', message=FALSE, warning=FALSE, echo = FALSE}

# caculate top10 fam
topfam <- asvs %>% 
  left_join(taxonomy) %>% select(sample, phylum, family) %>% unique() %>% 
  group_by(phylum, family) %>% summarise(Nsamples = n()) %>% arrange(desc(Nsamples)) %>% ungroup() %>%  
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    inner_join(taxonomy, by = 'asv') %>%
  group_by(phylum, family) %>% 
  summarise(reads = sum(count)) )%>% 
  filter(Nsamples == 109) %>%
  arrange(desc(reads)) %>% mutate(Order  = 1:nrow(.))

asvs %>%
  inner_join(taxonomy, by = 'asv') %>%
  left_join(topfam %>% transmute(family, toppf = family), by = 'family') %>%
  filter(!is.na(toppf)) %>%
  group_by(sample, toppf) %>%
  summarise(relab = sum(relab)) %>%
  rename(family = toppf) %>%
  ungroup() %>% 
  complete(sample, nesting(family), fill = list(relab = 0)) %>% 
  left_join(metadata %>% select(sample, Population)) %>%
  left_join(topfam) %>%
  mutate(familyp = paste(phylum, family)) %>%
  arrange(phylum) %>%
  ggplot(aes(x = familyp, y= relab)) +
  labs(y = " log Relative abundance", x ="") +
  geom_boxplot(aes(x = familyp, y= log(relab), fill=Population),
               position = position_dodge(width = .7))+
  #geom_point(aes(x = phylum, y= relab, col = Population, group = Population, alpha =0.1), position = position_dodge(width = .5)) +
  scale_fill_manual(values = pal) +
  theme(axis.title = element_text(size = 20, face = "bold"),
        legend.title = element_blank(),
        legend.position = "top",
        legend.direction = "horizontal",
        legend.text = element_text(size =20),
        axis.text.y = element_text(size =15),
        axis.text.x = element_text(size =15),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white'),
        panel.grid = element_blank(),
        legend.key=element_blank()) +
  coord_flip()+
  guides(alpha =FALSE)

```


<p>&nbsp;</p>


`r figr('tablefam', T, type = 'Table')`  **%  of reads of the top families with 100% prevalence of the reads of the phylum.**
```{r tablefam, echo=FALSE, message=FALSE, warnings=FALSE, echo = FALSE}
table <- asvs %>% 
  left_join(taxonomy) %>% select(sample, phylum, family) %>% unique() %>% 
  group_by(phylum, family) %>% summarise(Nsamples = n()) %>% arrange(desc(Nsamples)) %>% ungroup() %>%  
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    inner_join(taxonomy, by = 'asv') %>%
  group_by(phylum, family) %>% 
  summarise(reads = sum(count)) )%>% 
  arrange(desc(reads))%>% 
  left_join(
  asvs %>% filter(!is.na(count)) %>% 
    left_join(taxonomy, by = 'asv') %>%
    group_by(phylum) %>% 
    summarise(readsP = sum(count))
  ) %>%
  mutate(perc_readsP = reads/readsP)

table$perc_readsP <- formattable::percent(table$perc_readsP, digits = 0) 
table %>% na_if( 0) %>% select(-readsP) %>% kable()
```

<p>&nbsp;</p>


## F/B-ratio


```{r FBratiomasindex, fig.cap='**Figure 5.** A) Firmicutes Bacteriodetes ratio and B) BMI of wild and captive animals.', fig.height=4, fig.width=12, message=FALSE, warning=FALSE,  echo=FALSE}
ggarrange(
asvs %>%
  inner_join(taxonomy %>% select(asv, phylum), by = 'asv') %>%
  filter(phylum %in% c("Bacteroidetes", "Firmicutes")) %>%
  group_by(sample, phylum) %>%
  summarise(relab = sum(relab)) %>%
  ungroup() %>% 
  pivot_wider(names_from = phylum, values_from = relab) %>%
  mutate(FBratio = Firmicutes/ Bacteroidetes) %>%
  left_join(metadata %>% select(sample, Population, massindex)) %>%
  ggplot(aes(x = Population, y=  FBratio)) +
  geom_boxplot(aes(fill = Population)) +
  scale_fill_manual(values = pal) +
  ylab("F/B-ratio") +
  ylim(0,6) +
  ggtitle("A") +
  theme(axis.title.y = element_text(size = 20, face = "bold"),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size =20),
        axis.text.y = element_text(size =15),
        axis.text.x = element_text(size =15),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white'),
        panel.grid = element_blank(),
        legend.key=element_blank()),
metadata %>%
  ggplot(aes(x = Population, y=  massindex)) +
  geom_boxplot(aes(fill = Population)) +
  scale_fill_manual(values = pal) +
  ggtitle("B") +
  ylab("Body mass index") +
  theme(axis.title.y = element_text(size = 20, face = "bold"),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size =20),
        axis.text.y = element_text(size =15),
        axis.text.x = element_text(size =15),
        plot.title = element_text(size=22, face ="bold"),
        panel.background=element_rect(fill='white', colour='black'),
        strip.background=element_rect(fill='white', colour='white'),
        panel.grid = element_blank(),
        legend.key=element_blank()), common.legend = TRUE, ncol = 2)
```
Note: One outlier F/B from wild removed for visualization. F/B ratio One captive animal had no tooth width and therefore missing data on body mass index.


<p>&nbsp;</p>



`r figr('FBratiolmer', T, type = 'Table')`. LMM F/B-ratio table results.
```{r FBratiolmer, echo=FALSE, message=FALSE, warning=FALSE}
fb_data <- asvs %>%
  inner_join(taxonomy %>% select(asv, phylum), by = 'asv') %>%
  filter(phylum %in% c("Bacteroidetes", "Firmicutes")) %>%
  group_by(sample, phylum) %>%
  summarise(relab = sum(relab)) %>%
  ungroup() %>% 
  pivot_wider(names_from = phylum, values_from = relab) %>%
  mutate(FBratio = (Firmicutes/Bacteroidetes)) %>%
  left_join(metadata %>% select(sample, Population, Plate_No, NbReads))

mcheck<-function(obj,...){
  rs<-resid(obj)
  fv<-fitted(obj)
  par(mfrow=c(1,3))
  plot(fv,rs,xlab="FITTED VALUES",ylab="RESIDUALS")
  abline(h=0,lty=2,lwd=2)
  qqnorm(rs,xlab="NORMAL SCORES",ylab="ORDERED RESIDUALS",main="")
  qqline(rs,lty=2,lwd=2)
  hist(resid(obj),xlab="RESIDUALS",ylab="FREQUENCY",main="")
  par(mfrow=c(1,1))
  invisible(NULL)
}

# mead and sd per pop
#fb_data %>% group_by(Population) %>% summarise(mean = mean(FBratio),
#                                               sd = sd(FBratio) )
  
# FB ratio
m1 <- lmer(log(FBratio) ~ Population + (1|Plate_No), data = fb_data )
#summary(m1)
tab_model(m1, show.ci = 0.95, show.se = 0.95)
#tab_model(m1, show.ci = 0.95, show.se = 0.95, file = "../tables/FBratio.doc")
#mcheck(m1)
```


<p>&nbsp;</p>


## Body mass index

`r figr('massindexlm', T, type = 'Table')`. Lm body mass index table results. 
```{r massindexlm, echo=FALSE, message=FALSE, warning=FALSE}
# BMI
m2 <- lm(log(massindex) ~ Population, data = metadata %>% filter(!is.na(massindex))) # one captive without without
#summary(m2)
tab_model(m2, show.ci = 0.95, show.se = 0.95)
#mcheck(m2)
#tab_model(m2, show.ci = 0.95, show.se = 0.95, file = "../tables/massindex.doc")
```

<p>&nbsp;</p>

# Shared and unique ASVs

```{r VenndiagramAll, eval=FALSE, fig.cap='**Figure 6. Venndiagram all asvs. **', fig.height=5, fig.width=5, message=FALSE, warning=FALSE, include=FALSE}
Captive <- asvs %>% left_join(metadata) %>% filter(Population =="Captive") %>% select(asv) %>% unique() 
Captive <- Captive$asv
Wild <- asvs %>% left_join(metadata) %>% filter(Population =="Wild") %>% select(asv) %>% unique() 
Wild <- Wild$asv

venn.diagram(
        x = list(Captive, Wild),
        category.names = c("Captive" , "Wild"),
        filename = NULL,
        output=TRUE,
        
        # Output features
        imagetype="tiff" ,
        height = 480 , 
        width = 480 , 
        resolution = 300,
        compression = "lzw",
        
        # Circles
        lwd = 2,
        lty = 'blank',
        fill = pal,
        
        # Numbers
        cex = .6,
        fontface = "bold",
        
        # Set names
        cat.cex = 0.6,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.pos = c(-27, 27),
        cat.dist = c(0.055, 0.055)
)

```


<p>&nbsp;</p>

`r figr('shared-asvs-taxa', T, type = 'Table')`. Table of ASVs shared between wild and captive animals and their phylum assignment, sum of their count of reads and proportion of complete dataset.
```{r shared-asvs-taxa, message=FALSE, warning=FALSE}
# what taxa are the shared ASVs dominated by?
sharedASVs <- asvs %>% left_join(metadata %>% select(sample, Population)) %>% select(sample, asv, Population) %>%
  mutate(N = 1 ) %>% select (-sample) %>% unique() %>% pivot_wider(names_from = 'Population', values_from = 'N', values_fill = 0) %>%
  filter(Captive > 0 & Wild > 0) %>% select(asv)

# sum by phylum
t <- sharedASVs %>%
  left_join(taxonomy) %>%
  left_join(topphyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
  replace_na(list('topph' = 'Other phyla')) %>%
  group_by(topph) %>%
  summarise(NAsvs = n()) %>% rename(phylum = topph) %>%
  left_join(
    sharedASVs %>%
    left_join(asvs) %>%
    left_join(taxonomy) %>%
    left_join(topphyla %>% transmute(phylum, topph = phylum), by = 'phylum') %>%
    replace_na(list('topph' = 'Other phyla')) %>%
    group_by(topph) %>%
    summarise(sumReads = sum(count)) %>% 
    mutate(prop_reads = sumReads/(asvs %>% summarise(sumcount = sum(count)) %>% pull())) %>%
    rename(phylum = topph)) %>% # follow the distribution of phylum full dataset)
  left_join(topphyla %>% select(phylum, Order)) %>% #join in to sort by 
    replace_na(list('Order' = 8)) %>% 
    arrange(Order) %>% select(-Order)

t %>% kable()

# write table
#t %>% mutate_if(is.numeric, ~round(., 3)) %>%
#  write.table(file ="Shared.txt", row.names = F)
```


<p>&nbsp;</p>

So, how many sequences belonged to the shared ASVs? `r sharedASVs %>% left_join(asvs) %>% summarise(sumcount = sum(count)) %>% pull()`, this is equal to `r sharedASVs %>% left_join(asvs) %>% summarise(sumcount = sum(count)) %>% pull() / asvs %>% summarise(sumcount = sum(count))*100`% of the complete data!


<p>&nbsp;</p>

# R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in
`r figr('cite-packages', T, type = 'Table')`.


```{r cite-packages, cache = FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```



<p>&nbsp;</p>

# References